---
title: "Opioid, ED Visit, & Mortality"
output: html_notebook
---



#1. Prepare Data
## 1.1 Activate Packages
```{r}
#install.packages("knitr")
library(dplyr)
library(readr)
library(lubridate)
library(tidyr)
library(stringr)
library(janitor)
library(sdmTMB)
library(tigris)
library(sf)
library(here)
library(ggplot2)
library(dplyr)
library(knitr)
library(psych)

install.packages("patchwork")
library(patchwork)
```

## 1.2 Merge Datasets
```{r figure2, echo=FALSE, message=FALSE, warning=FALSE}
###############################

#repos <- "/Users/yuihuiyao/Library/CloudStorage/Box-Box/2025-5-ER-Spatial-Autocorrelation"
#repos <- "C:/Users/julia/Box/Personal/Research/2025-5-ER-Spatial-Autocorrelation"

repos <- here()

file_path <- paste(repos, "/Data/ER_opioid_2016_2019_cleaned.csv", sep = "")
if (!file.exists(file_path)) {
  stop("File not found: Data/ER_opioid_2016_2019_cleaned.csv")
}
er_data <- read_csv(file_path, show_col_types = FALSE)

# Keep a subset of North / Central AL (as in your script)
keep_cnty <- toupper(c(
  "Lauderdale","Limestone","Madison","Jackson","Colbert","Franklin",
  "Lawrence","Morgan","Marshall","Dekalb","Marion","Winston",
  "Cullman","Blount","Etowah","Cherokee","Lamar","Fayette",
  "Walker","Jefferson","St. Clair","Pickens","Tuscaloosa","Shelby"
))

er_data <- er_data |>
  mutate(County = toupper(County),
         Date = as.Date(Date)) |>
  filter(County %in% keep_cnty)


# Read Data_ARCOS and clean for next process
Data_ARCOS <- read_csv(paste(repos, "/BigData/ARCOS_AL_2016_2019_cleaned.csv", sep = ""))

Data_ARCOS_filted <- Data_ARCOS%>%
  filter(drug %in% c("OXYCODONE", "HYDROCODONE", "BUPRENORPHINE", "METHADONE"))%>%
  filter(county %in% toupper(c(
    "Lauderdale", "Limestone", "Madison", "Jackson", "Colbert", "Franklin",
    "Lawrence", "Morgan", "Marshall", "Dekalb", "Marion", "Winston",
    "Cullman", "Blount", "Etowah", "Cherokee", "Lamar", "Fayette",
    "Walker", "Jefferson", "St. Clair", "Pickens", "Tuscaloosa", "Shelby"
  ))
  )%>%
  #mutate(mme = mme/1000000)%>%
  mutate(Date = mdy(date))%>%
  mutate(County = toupper(county))

arcos_wide_mme <- Data_ARCOS_filted %>%
  pivot_wider(
    id_cols    = c(County, Date),
    names_from = drug,
    values_from= mme,
    values_fill = 0
  )

df_er_drug <- left_join(
  er_data,
  arcos_wide_mme,
  by = c("County", "Date")
)

###############################

# Presence at daily level
df_er_drug <- df_er_drug |>
  mutate(
    year = year(Date),
    presence = if_else(Count > 0, 1L, 0L)
  )

# Aggregate to county × year
df_er_drug_annual <- df_er_drug %>%
  group_by(County, year) %>%
  summarise(
    # population for the year (choose one: mean/first/median)
    pop_year = first(pop),

    # ER annual
    ER_annual = sum(Count, na.rm = TRUE),

    # Drug annual
    OXYCODONE_annual     = sum(OXYCODONE,     na.rm = TRUE),
    HYDROCODONE_annual   = sum(HYDROCODONE,   na.rm = TRUE),
    BUPRENORPHINE_annual = sum(BUPRENORPHINE, na.rm = TRUE),
    METHADONE_annual     = sum(METHADONE,     na.rm = TRUE),

    presence = as.integer(sum(Count, na.rm = TRUE) > 0),
    .groups = "drop"
  ) %>%
  mutate(
    # per-capita (proportion) and per 100k (common epidemiology scale)
    ER_annual_percapita           = ER_annual / pop_year,
    ER_annual_per100k             = ER_annual_percapita * 1e5,

    OXYCODONE_annual_percapita    = OXYCODONE_annual     / pop_year,
    OXYCODONE_annual_per100k      = OXYCODONE_annual_percapita * 1e5,

    HYDROCODONE_annual_percapita  = HYDROCODONE_annual   / pop_year,
    HYDROCODONE_annual_per100k    = HYDROCODONE_annual_percapita * 1e5,

    BUPRENORPHINE_annual_percapita = BUPRENORPHINE_annual / pop_year,
    BUPRENORPHINE_annual_per100k   = BUPRENORPHINE_annual_percapita * 1e5,

    METHADONE_annual_percapita    = METHADONE_annual     / pop_year,
    METHADONE_annual_per100k      = METHADONE_annual_percapita * 1e5
  )

###############################

mortality <- read_csv(paste(repos, "/Data/CDCWonder Opioid Related Mortality AL 2016-2019.csv", sep = ""))

mortality_filted <- mortality%>%
  mutate(County = toupper(gsub(" County, AL", "", County)))%>%
  filter(County %in% toupper(c(
    "Lauderdale", "Limestone", "Madison", "Jackson", "Colbert", "Franklin",
    "Lawrence", "Morgan", "Marshall", "Dekalb", "Marion", "Winston",
    "Cullman", "Blount", "Etowah", "Cherokee", "Lamar", "Fayette",
    "Walker", "Jefferson", "St. Clair", "Pickens", "Tuscaloosa", "Shelby"
  )))%>%
  mutate(
    Deaths = suppressWarnings(as.numeric(Deaths)),      # "Suppressed" → NA
    CrudeRate = suppressWarnings(as.numeric(CrudeRate)),
    deaths_per100k = Deaths / Population * 1e5
  )%>%
  rename(year = Year,
         pop_CDC=Population)


# Merge
df_er_drug_mortality <- left_join(
  df_er_drug_annual,
  mortality_filted,
  by = c("County", "year")
)

###############################

acs <- read_csv(paste(repos, "/Data/ACS 2019 Data Subset.csv", sep = ""))



# 1) Build a safe, readable column name from Concept + Label
clean_colname <- function(concept, label) {
  # drop leading "Estimate!!"
  label2 <- str_remove(label, "^Estimate!!")
  # collapse "!!" / ":" and other punctuation to separators
  label2 <- str_replace_all(label2, "!!", "_")
  label2 <- str_replace_all(label2, "[:;,-]+", "")
  # build combined name, then snake case + de-duplicate
  nm <- paste(concept, label2, sep = "__")
  # nm <- janitor::make_clean_names(nm)
  nm
}

acs2 <- acs %>%
  mutate(
    colname = clean_colname(Concept, Label)
  )

# 2) If there are duplicates per (State, County, Full_FIPS, colname), choose the first non-NA
acs_collapsed <- acs2 %>%
  group_by(State, County, Full_FIPS, colname) %>%
  summarise(Value = dplyr::first(na.omit(Value)), .groups = "drop")


# 3) Pivot to wide
acs_wide <- acs_collapsed %>%
  pivot_wider(
    id_cols    = c(State, County, Full_FIPS),
    names_from = colname,
    values_from = Value,
    values_fill = NA_real_
  ) %>%
  clean_names()%>%
  mutate(CountyCode = full_fips)# final pass to ensure valid names

# Merge three imputated mortality
df_er_drug_mortality_acs <- left_join(
  df_er_drug_mortality,
  acs_wide,
  by = c("CountyCode")
  )
#summary(mortality_filted)

###############################

df_er_drug_mortality_acs <- df_er_drug_mortality_acs%>%
  mutate(
    unemployee_rate = employment_status_for_the_population_16_years_and_over_total_in_labor_force_civilian_labor_force_unemployed/employment_status_for_the_population_16_years_and_over_total_in_labor_force_civilian_labor_force,

    gini_index = gini_index_of_income_inequality_gini_index,

    median_income = median_household_income_in_the_past_12_months_in_2021_inflation_adjusted_dollars_median_household_income_in_the_past_12_months_in_2021_inflationadjusted_dollars,

    poverty_rate = poverty_status_in_the_past_12_months_by_sex_by_age_total_income_in_the_past_12_months_below_poverty_level/poverty_status_in_the_past_12_months_by_sex_by_age_total,

    # sum all columns about disability
    disability = rowSums(
      across(starts_with("sex_by_age_by_disability"),
             ~ as.numeric(.x)),           # ensure numeric
      na.rm = TRUE),
    # (optional) per-capita / per-100k versions
    disability_rate      = disability / total_population_total,
    disability_per100k   = disability_rate * 1e5,

    # sum all columns about noinsurance
    noinsurance = rowSums(
      across(starts_with("types_of_health_insurance_coverage"),
             ~ replace_na(as.numeric(.x), 0)),
      na.rm = TRUE
    ),
    # Optional: per-capita or per-100k version
    noinsurance_rate    = noinsurance / total_population_total,
    noinsurance_per100k = noinsurance_rate * 1e5)

###############################

df_select <- df_er_drug_mortality_acs%>%
  select(
    County,
    year,
    pop_CDC,
    # Deaths
    Deaths,
    # ER visit
    ER_annual_percapita,
    # Drugs
    OXYCODONE_annual_percapita, HYDROCODONE_annual_percapita,   BUPRENORPHINE_annual_percapita, METHADONE_annual_percapita,

    # ACS
    unemployee_rate,    gini_index, poverty_rate,   disability_rate, noinsurance_rate
    )

###############################

scale_vars <-  c(
    # Deaths
    "Deaths",
    # ER visit
    "ER_annual_percapita",
    # Drugs
    "OXYCODONE_annual_percapita",   "HYDROCODONE_annual_percapita", "BUPRENORPHINE_annual_percapita", "METHADONE_annual_percapita",

    # ACS
    "unemployee_rate",  "gini_index",   "poverty_rate", "disability_rate", "noinsurance_rate"
    )

df_select_z <- df_select %>%
  mutate(Deaths_rate= Deaths / pop_CDC)%>%
  mutate(across(all_of(scale_vars), ~ as.numeric(scale(.x)), .names = "{.col}_z"))


###############################
# initialization
###############################

###############################

df_select_z$time_index <- df_select_z$year - 2018
df_select_z$intervention2017 <- as.numeric(df_select_z$year >= 2018)
df_select_z_int <- df_select_z
#df_select_z_int$Deaths[is.na(df_select_z_int$Deaths)] <- (9+1)/2

df_select_z_int$rural <- 0

df_select_z_int$rural[which(df_select_z_int$County == "BLOUNT")] <- 2
df_select_z_int$rural[which(df_select_z_int$County == "CHEROKEE")] <- 6
df_select_z_int$rural[which(df_select_z_int$County == "COLBERT")] <- 4
df_select_z_int$rural[which(df_select_z_int$County == "CULLMAN")] <- 5
df_select_z_int$rural[which(df_select_z_int$County == "ETOWAH")] <- 4
df_select_z_int$rural[which(df_select_z_int$County == "FAYETTE")] <- 6
df_select_z_int$rural[which(df_select_z_int$County == "FRANKLIN")] <- 5
df_select_z_int$rural[which(df_select_z_int$County == "JACKSON")] <- 5
df_select_z_int$rural[which(df_select_z_int$County == "JEFFERSON")] <- 2
df_select_z_int$rural[which(df_select_z_int$County == "LAMAR")] <- 6
df_select_z_int$rural[which(df_select_z_int$County == "LAUDERDALE")] <- 4
df_select_z_int$rural[which(df_select_z_int$County == "LAWRENCE")] <- 4
df_select_z_int$rural[which(df_select_z_int$County == "LIMESTONE")] <- 3
df_select_z_int$rural[which(df_select_z_int$County == "MADISON")] <- 3
df_select_z_int$rural[which(df_select_z_int$County == "MARION")] <- 6
df_select_z_int$rural[which(df_select_z_int$County == "MARSHALL")] <- 5
df_select_z_int$rural[which(df_select_z_int$County == "MORGAN")] <- 4
df_select_z_int$rural[which(df_select_z_int$County == "PICKENS")] <- 3
df_select_z_int$rural[which(df_select_z_int$County == "SHELBY")] <- 2
df_select_z_int$rural[which(df_select_z_int$County == "TUSCALOOSA")] <- 3
df_select_z_int$rural[which(df_select_z_int$County == "WALKER")] <- 2
df_select_z_int$rural[which(df_select_z_int$County == "WINSTON")] <- 6

df_select_z_int$rural_factor <- factor(df_select_z_int$rural)

###############################


counties_sf <- tigris::counties(state = "AL", cb = TRUE, year = 2023) |>
  st_transform(26916) |>
  mutate(County = toupper(NAME))

coords_m  <- st_coordinates(st_centroid(st_geometry(counties_sf)))
county_xy <- counties_sf |>
  st_drop_geometry() |>
  transmute(
    County,
    X = coords_m[, "X"] / 1000,  # km
    Y = coords_m[, "Y"] / 1000
  )

# Merge coords into model frame
df <- df_select_z_int %>%
  left_join(county_xy, by = "County") %>%
  filter(is.finite(X), is.finite(Y), is.finite(pop_CDC))%>%
  mutate(time_index = match(year, sort(unique(year))),
         log_pop    = log(pop_CDC))

mesh <- make_mesh(
  df,
  xy_cols = c("X","Y"),
  cutoff = diff(range(df$X, na.rm = TRUE)) / 10
)

###############################

df$time_index_factor <- factor(df$time_index)

###############################
# County-level summary for Study Population / Table 1 (from ACS-merged data)
county_chars <- df_er_drug_mortality_acs %>%
  group_by(County) %>%
  summarise(
    population    = round(mean(pop_CDC, na.rm = TRUE)),
    median_income = round(median(median_income, na.rm = TRUE), -2),
    pct_uninsured = round(100 * mean(noinsurance_rate, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  mutate(
    County_display = str_to_title(County),
    HPSA = NA_integer_   # HPSA not in current data
  )

###############################

write.csv(df, paste0(repos,"/Data/all.csv"))

```

## 1.3 Read Merged Data [Optional]
```{r}
#df <- read.csv(paste0(repos,"/Data/all.csv"))
```


# 2. Descriptive Statistics

##  [[[Table 1 (Part 1)]]]. County Characteristics in Northwestern Alabama, 2016–2019

```{r table1-study-population, results='asis'}
county_chars <- df %>%
  mutate(County = str_to_title(County)) %>%
  group_by(County) %>%
  summarise(
    population    = round(mean(as.numeric(pop_CDC), na.rm = TRUE)),
    pct_gini = round(100 * mean(as.numeric(gini_index), na.rm = TRUE), 1),
    pct_uninsured = round(100 * mean(as.numeric(noinsurance_rate), na.rm = TRUE), 1),
    pct_unemployee = round(100 * mean(as.numeric(unemployee_rate), na.rm = TRUE), 1),
    pct_poverty = round(100 * mean(as.numeric(poverty_rate), na.rm = TRUE), 1),
    pct_disability = round(100 * mean(as.numeric(disability_rate), na.rm = TRUE), 1),
    median_income = NA_real_,  # not in all.csv
    HPSA         = NA_integer_,
    .groups = "drop"
  )

county_chars

#write.csv(county_chars, paste0(repos,"/Data/table1.csv"))

```



## [[[Table 1 (Part 2)]]]. County Characteristics in Northwestern Alabama, 2016–2019

```{r table2-ed-deaths, results='asis'}
# Regional totals and by-year summary from df (all.csv)
df_num <- df %>%
  mutate(
    pop_CDC   = as.numeric(pop_CDC),
    Deaths    = as.numeric(Deaths),
    ER_percap = as.numeric(ER_annual_percapita)
  )%>%
  mutate(
    ed_per_10k   = round(ER_percap* 1e4, 2),
    deaths_per_100k = round(Deaths / pop_CDC * 1e5, 2)
  )

ed_death_wide <- df_num %>%
  # Select only the columns needed for the final table
  select(County, rural, year, ed_per_10k, deaths_per_100k) %>%
  # Pivot to create year-based subcolumns
  pivot_wider(
    names_from = year,
    values_from = c(ed_per_10k, deaths_per_100k),
    names_glue = "{year}_{.value}"
  ) 


# %>%
#   # Optional: Reorder columns so the years are grouped together
#   select(County, sort(peek_vars()))

write.csv(ed_death_wide, paste0(repos,"/Data/table2.csv"))


```


## [[[Figure 1]]]. County-Level Emergency Department Visit Rates, 2016–2019s
```{r}
fig1_data <- df_num %>%
  filter(year >= 2016, year <= 2019) %>%
  mutate(
    year = as.numeric(as.character(year))
  )

ggplot(
  fig1_data,
  aes(x = year, y = ed_per_10k, group = 1)
) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.2) +
  facet_wrap(~ County, scales = "free_y", ncol = 3) +
  scale_x_continuous(
    breaks = c(2016, 2017, 2018, 2019)
  ) +
  labs(
    title = "County-Level Emergency Department Visit Rates, 2016–2019",
    x = "Year",
    y = "ED Visits per 10,000 Population"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```


# 3. Imputation and Model Comparison Leave-One-Out Cross-Validatio with Truncated Expectation

### [!!! Please skip 3.0], we use selected model from Yao's Output

## 3.0 Comparison All Models [Optional]
```{r}
getFit <- function(df, df_select_z_int, formula, family = poisson(link = "log"), spatial = "off", spatiotemporal = "off") {

  tr_ex <- function(mu, lower = 1, upper = 9) {
    mu_tr <- sum(((lower:upper) * dpois(lower:upper, mu))) / (ppois(upper, mu) - ppois(lower - 1, mu))
    mu_tr <- ifelse(mu_tr > 9, 9, mu_tr)
    mu_tr <- ifelse(mu_tr < 1, 1, mu_tr)
  }

  tr_ex <- Vectorize(tr_ex, "mu")

  df$Deaths[is.na(df_select_z_int$Deaths)] <- (9+1)/2

  ###############################


  log_lik_old <- -Inf
  log_lik_new <- Inf
  diff_lik <- Inf
  si <- -1
  tol <- 1e-4

  r <- 0

  previous_fit <- NULL

    mesh <- make_mesh(
      df,
      xy_cols = c("X","Y"),
      cutoff = diff(range(df$X, na.rm = TRUE)) / 10
    )

  while ((diff_lik > tol) & (si == -1)) {
    r <- r + 1

    cat("r:", r, ", log_lik:", log_lik_old, "\n")
    fit <- try(sdmTMB(
      formula = formula,
      data = df,
      mesh = mesh,
      time = "time_index",
      family  = family,
      spatial = spatial,
      spatiotemporal = spatiotemporal,
      offset  = df$log_pop,
      previous_fit = previous_fit
    ), silent = TRUE)

    if (class(fit)[1] == "try-error") {
      cat("  sdmTMB fitting failed at iteration", r, "\n")
      break
    }

    previous_fit <- fit

    if (r == 1) {
      init_fit <- previous_fit
    }

    log_lik_new <- logLik(fit)

    fitted_value <- fitted(fit)

    tmp_fitted_value <- fitted_value[is.na(df_select_z_int$Deaths)]
    tmp_fitted_value <- tr_ex(tmp_fitted_value)

    df$Deaths[is.na(df_select_z_int$Deaths)] <- tmp_fitted_value

    diff_lik <- (log_lik_old - log_lik_new)
    si <- sign(log_lik_old - log_lik_new)
    diff_lik <- abs(diff_lik)

    log_lik_old <- log_lik_new

  }

  list("init_fit" = init_fit, "fit" = previous_fit, "imputed_df" = df)

}




cv_getFit_loo <- function(df, df_select_z_int, formula,
                          family = poisson(link = "log"), spatial = "off", spatiotemporal = "off") {

  n <- nrow(df)

  notna <- which(!is.na(df$Deaths))
  K <- length(notna)

  cv_st <- matrix(NA_real_, nrow = K, ncol = 4)
  colnames(cv_st) <- c("IM_RMSE", "IM_MAE","EM_RMSE", "EM_MAE")

  r <- 0
  for (k in notna) {
    r <- r + 1
    cat("  Fold", r, "of", K, "\n")
    test_idx  <- k
    train_idx <- 1:n
    train_idx <- train_idx[train_idx != test_idx]

    df_train        <- df[train_idx, , drop = FALSE]
    df_select_train <- df_select_z_int[train_idx, , drop = FALSE]
    df_test         <- df[test_idx, , drop = FALSE]

    # Fit model on training data
      fit_k <- try(getFit(df_train, df_select_train, formula,
                          family, spatial = spatial, spatiotemporal = spatiotemporal), silent = TRUE)


    if (class(fit_k)[1] == "try-error") {
      cv_st[r, "IM_RMSE"] <- NA
      cv_st[r, "IM_MAE"] <- NA
      cv_st[r, "EM_RMSE"] <- NA
      cv_st[r, "EM_MAE"] <- NA
      next
    }

    # Predict on test data
    pred_init <- try(predict(fit_k$init_fit, newdata = df_test, offset = df_test$log_pop, type = "response"), silent = TRUE)

    if (class(pred_init)[1] == "try-error") {
      cv_st[r, "IM_RMSE"] <- NA
      cv_st[r, "IM_MAE"] <- NA
    } else {
      pred_init <- pred_init$est
      # Use only rows with observed Deaths
      pred_notna    <- pred_init[!is.na(df_test$Deaths)]
      deaths_notna  <- df_test$Deaths[!is.na(df_test$Deaths)]

      rmse <- (((pred_notna - deaths_notna)^2))
      mae <- (abs(pred_notna - deaths_notna))
      cv_st[r, "IM_RMSE"] <- rmse
      cv_st[r, "IM_MAE"] <- mae
    }



    pred <- try(predict(fit_k$fit, newdata = df_test, offset = df_test$log_pop, type = "response"), silent = TRUE)

    if (class(pred)[1] == "try-error") {
      cv_st[r, "EM_RMSE"] <- NA
      cv_st[r, "EM_MAE"] <- NA
    } else {
      pred <- pred$est
      # Use only rows with observed Deaths
      pred_notna    <- pred[!is.na(df_test$Deaths)]
      deaths_notna  <- df_test$Deaths[!is.na(df_test$Deaths)]

      rmse <- (((pred_notna - deaths_notna)^2))
      mae <- (abs(pred_notna - deaths_notna))
      cv_st[r, "EM_RMSE"] <- rmse
      cv_st[r, "EM_MAE"] <- mae
    }


  }

  cv_st <- colMeans(cv_st, na.rm = TRUE)
  cv_st[c(1, 3)] <- sqrt(cv_st[c(1, 3)])
  cv_st
}


###############################

###############################

cor_df <- df %>% select(
  ER_annual_percapita_z,
  OXYCODONE_annual_percapita_z,
  HYDROCODONE_annual_percapita_z,
  BUPRENORPHINE_annual_percapita_z,
  METHADONE_annual_percapita_z,
  unemployee_rate_z,
  gini_index_z,
  poverty_rate_z,
  disability_rate_z,
  noinsurance_rate_z
)

corrplot::corrplot(cor(cor_df, method = "spearman"))

###############################

full_formula <- Deaths ~ ER_annual_percapita_z +
  OXYCODONE_annual_percapita_z +
  #HYDROCODONE_annual_percapita_z +
  BUPRENORPHINE_annual_percapita_z +
  METHADONE_annual_percapita_z +
  #unemployee_rate_z +
  #gini_index_z +
  poverty_rate_z +
  disability_rate_z +
  noinsurance_rate_z +
  rural_factor

fit0 <- getFit(df, df_select_z_int, full_formula)

car::vif(fit0$fit)

###############################
  #remove the highest VIF until all less than 5
###############################

wrapper <- function(X, formulas, df, df_select_z_int) {
  
  formula <- formulas[[X]]
  
  cv <- list()
  
  cv[[1]] <- try(cv_getFit_loo(df, df_select_z_int, formula, spatial = "off", spatiotemporal = "off"), silent = TRUE)
  cv[[2]] <- try(cv_getFit_loo(df, df_select_z_int, formula, spatial = "on", spatiotemporal = "off"), silent = TRUE)
  cv[[3]] <- try(cv_getFit_loo(df, df_select_z_int, formula, spatial = "off", spatiotemporal = "ar1"), silent = TRUE)
  cv[[4]] <- try(cv_getFit_loo(df, df_select_z_int, formula, spatial = "on", spatiotemporal = "ar1"), silent = TRUE)
  cv[[5]] <- try(cv_getFit_loo(df, df_select_z_int, formula, family = nbinom2(link = "log"), spatial = "off", spatiotemporal = "off"), silent = TRUE)
  cv[[6]] <- try(cv_getFit_loo(df, df_select_z_int, formula, family = nbinom2(link = "log"), spatial = "on", spatiotemporal = "off"), silent = TRUE)
  cv[[7]] <- try(cv_getFit_loo(df, df_select_z_int, formula, family = nbinom2(link = "log"), spatial = "off", spatiotemporal = "ar1"), silent = TRUE)
  cv[[8]] <- try(cv_getFit_loo(df, df_select_z_int, formula, family = nbinom2(link = "log"), spatial = "on", spatiotemporal = "ar1"), silent = TRUE)

  out <- NULL
  
  for (i in 1:4) {
    if (class(cv[[i]]) != "try-error") {
      out <- c(out, cv[[i]])
    } else {
      out <- c(out, rep(NA, 4))
    }
  }
  
  out
  
}



base_vars <- c(
  "ER_annual_percapita_z",
  "OXYCODONE_annual_percapita_z",
  "BUPRENORPHINE_annual_percapita_z",
  "METHADONE_annual_percapita_z",
  "rural_factor"
)

covars <- c("poverty_rate_z", "disability_rate_z", "noinsurance_rate_z", "rural_factor")

make_formula <- function(vars){
  rhs_main <- if(length(vars) > 0){
    paste0("(", paste(vars, collapse = " + "), ") * intervention2017")
  } else {
    "intervention2017"
  }
  rhs_cov <- paste(covars, collapse = " + ")
  as.formula(paste("Deaths ~", rhs_main, "+", rhs_cov))
}

make_formula1 <- function(vars){
  rhs_main <- if(length(vars) > 0){
    paste0("(", paste(vars, collapse = " + "), ") + intervention2017")
  } else {
    "intervention2017"
  }
  rhs_cov <- paste(covars, collapse = " + ")
  as.formula(paste("Deaths ~", rhs_main, "+", rhs_cov))
}

subsets <- unlist(
  lapply(0:length(base_vars), function(k)
    combn(base_vars, k, simplify = FALSE)
  ),
  recursive = FALSE
)

formulas <- NULL

formulas <- lapply(subsets, make_formula)
names(formulas) <- sapply(subsets, function(v)
  if(length(v) == 0) "none" else paste(v, collapse = "_")
)

formulas1 <- NULL

formulas1 <- lapply(subsets, make_formula1)
names(formulas) <- sapply(subsets, function(v)
  if(length(v) == 0) "none" else paste(v, collapse = "_")
)


formulas2 <- list()

for  (i in 1:length(formulas)) {
  formulas2[[i]] <- formulas[[i]]
  formulas2[[length(formulas) + (i)]] <- formulas1[[i]]
}


nn <- length(formulas2)

formulas2[[nn + 1]] <- Deaths ~ ER_annual_percapita_z
formulas2[[nn + 2]] <- Deaths ~ OXYCODONE_annual_percapita_z
formulas2[[nn + 3]] <- Deaths ~ BUPRENORPHINE_annual_percapita_z
formulas2[[nn + 4]] <- Deaths ~ METHADONE_annual_percapita_z
formulas2[[nn + 5]] <- Deaths ~ poverty_rate_z
formulas2[[nn + 6]] <- Deaths ~ disability_rate_z
formulas2[[nn + 7]] <- Deaths ~ noinsurance_rate_z
formulas2[[nn + 8]] <- Deaths ~ rural_factor
formulas2[[nn + 9]] <- Deaths ~ intervention2017


############################### Parallel

nform <- length(formulas2)

cl <- parallel::makeCluster(min(floor(18), nform))
parallel::clusterExport(cl, c("getFit", "cv_getFit_loo", "wrapper"))
parallel::clusterEvalQ(cl, library(dplyr))
parallel::clusterEvalQ(cl, library(readr))
parallel::clusterEvalQ(cl, library(lubridate))
parallel::clusterEvalQ(cl, library(tidyr))
parallel::clusterEvalQ(cl, library(stringr))
parallel::clusterEvalQ(cl, library(janitor))
parallel::clusterEvalQ(cl, library(sdmTMB))
parallel::clusterEvalQ(cl, library(tigris))
parallel::clusterEvalQ(cl, library(sf))


out <- parallel::parLapplyLB(cl, X = 1:nform, fun = wrapper, formulas = formulas2, df = df, df_select_z_int = df_select_z_int)

#save(out, file = paste0(repos, "/Data/Yao_out.Rdat"))

```

## 3.1 The Best Model (From Yao's output)
```{r}
load(paste0(repos, "/Data/Yao_out.Rdat"))
```


```{r}
outmatrix <- matrix(NA, nrow = length(out), ncol = length(out[[1]]))

for (i in 1:length(out)) {
  
  outmatrix[i, ] <- out[[i]]
  
  
}

# 1. Identify the index of the best performing model
# (Using your criteria: min value in specific columns)
best_idx <- which(outmatrix == min(outmatrix[, c(2, 4, 6, 8, 10, 12, 14, 16)]), arr.ind = TRUE)[1]

# 2. Extract that specific formula from your list
best_formula <- formulas2[[best_idx]]

# 3. Re-fit the model to get the full sdmTMB object
# Ensure 'spatial' is "on" to get the Omega (spatial random effects)
final_fit <- getFit(
  df = df, 
  df_select_z_int = df_select_z_int, 
  formula = best_formula, 
  spatial = "off", # "on"
  spatiotemporal = "off"
)

final_fit


print ("Imputed Data")
final_fit$imputed_df


```

```{r}
fit0$fit$data$Deaths

formula_imputated<- Deaths ~ ER_annual_percapita_z

fit_imputated <- sdmTMB(
      formula = formula_imputated,
      data = final_fit$imputed_df,
      mesh = mesh,

      offset  = df$log_pop
    )

fit_imputated   
```



## 3.2 Scatted plot: Relationship Between ED Visits and Mortality [Optional]
```{r}
# 1. Get the data
# We use fit0$imputed_df because getFit returns the df with filled values in that list element.
# If you want to plot a different model result, run: fit_best <- getFit(df, df_select_z_int, your_best_formula)
plot_data <- fit0$fit$data%>%
  mutate(Death_10k=Deaths*10000/pop_CDC)%>%
  mutate(ER_annual_10k=ER_annual_percapita*10000)
  

# 2. Identify Observed vs. Adjusted points
# We look at the original 'df_select_z_int' to see which Death values were missing (suppressed)
plot_data$Status <- ifelse(is.na(plot_data$Deaths_z), "Adjusted", "Observed")

# 3. Calculate Correlation (using Spearman as you did in the corrplot)
cor_val <- cor(plot_data$ER_annual_10k, plot_data$Death_10k, method = "spearman", use = "complete.obs")
cor_label <- paste("Spearman Cor:", round(cor_val, 3))

# 4. Generate the Scatter Plot
ggplot(plot_data, aes(x = ER_annual_10k, y = Death_10k)) +
  # Add points: distinct colors for Observed vs Adjusted
  geom_point(aes(color = Status), size = 2, alpha = 0.8) +
  
  # Define specific colors: Solid Blue for Observed, Light Blue for Adjusted
  scale_color_manual(values = c("Observed" = "darkblue", "Adjusted" = "lightblue")) +
  
  # # Add the correlation text label on the plot (top-left corner)
  # annotate("text", 
  #          x = min(plot_data$ER_annual_percapita_z, na.rm = TRUE), 
  #          y = max(plot_data$Deaths, na.rm = TRUE), 
  #          label = cor_label, 
  #          hjust = 0, vjust = 1, size = 5, fontface = "bold") +
  
  # Labels and clean theme
  labs(
    title = "Standardized ED Visits vs. Deaths",
    x = "ED Visits (per 10k)",
    y = "Deaths (per 10k)"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 11)
  )
```

## 3.3 Scatted plot: (Yearly) Relationship Between ED Visits and Mortality [Optional]
```{r}
# 4. Generate the Yearly Faceted Scatter Plot
ggplot(plot_data, aes(x = ER_annual_10k, y = Death_10k)) +
  # Add points: distinct colors for Observed vs Adjusted
  geom_point(aes(color = Status), size = 1.5, alpha = 0.6) +
  
  # Define specific colors: Solid Blue for Observed, Light Blue for Adjusted
  scale_color_manual(values = c("Observed" = "darkblue", "Adjusted" = "lightblue")) +
  
  # Partition the plot by Year
  # scales = "fixed" is standard for comparison, 
  # but use "free" if the ranges differ drastically across years
  facet_wrap(~ year) + 
  
  # Labels and clean theme
  labs(
    title = "Evolution of Standardized ED Visits vs. Deaths",
    subtitle = "Comparing Observed and Adjusted Metrics across Years",
    x = "ED Visits (per 10k)",
    y = "Deaths (per 10k)"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 11),
    strip.background = element_rect(fill = "white"), # Clean facet headers
    strip.text = element_text(face = "bold")
  )
```

## 3.4 Scatted plot with lines (seperated by imputated and raw data): (Yearly) Relationship Between ED Visits and Mortality [Optional]
```{r}
# 1. Construct the plot with aligned labels
p <- ggplot(plot_data, aes(x = ER_annual_10k, y = Death_10k, color = Status, fill = Status)) +
  geom_point(size = 1, alpha = 0.2) +
  
  # 'group = Status' ensures smooth, straight vectors for each category
  geom_smooth(aes(group = Status), 
              method = "lm", 
              formula = y ~ x, 
              se = TRUE, 
              alpha = 0.1, 
              linewidth = 0.7) + 
  
  # High-Contrast Palette: Deep Navy and Vibrant Orange
  scale_color_manual(values = c("Observed" = "#002D62", "Adjusted" = "#D35400")) +
  scale_fill_manual(values = c("Observed" = "#002D62", "Adjusted" = "#D35400")) +
  
  # Aligning labels to match Trend_combined.png
  labs(
    x = "Emergency Department Visits (per 10k)",
    y = "Opioid-Related Mortality (per 10k)",
    color = "Status",
    fill = "Status"
  ) +
  
  facet_wrap(~ year) + 
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )


# 2. Save the high-resolution artifact
ggsave("Opioid_Trend_Analysis.png", plot = p, width = 6, height = 4.2, dpi = 300)

p
```

## 3.5 Scatted plot with one combined line: (Yearly) Relationship Between ED Visits and Mortality [Optional]
```{r}
# 1. Construct the Unified Visualization
# By removing 'color = Status', we treat the population as a single 'mind in motion'
p2 <- ggplot(plot_data, aes(x = ER_annual_10k, y = Death_10k)) +
  
  # Layer 1: The 'Community Exposure' (Individual Points)
  # Set to black with a low alpha to highlight the density of the lattice
  geom_point(size = 1, alpha = 0.2, color = "black") +
  
  # Layer 2: The 'Whole' Regression Line
  # This provides a single, straight mathematical vector for each facet
  geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    se = TRUE, 
    color = "black",  # Forces the line to black
    fill = "black",   # Sets the confidence interval to a subtle grey
    alpha = 0.1, 
    linewidth = 0.7
  ) + 
  
  # Layer 3: Temporal Faceting and Structural Refinement
  facet_wrap(~ year) + 
  theme_bw() +
  theme(
    legend.position = "none",  # Legend is no longer necessary for a unified plot
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "Emergency Department Visits (per 10k)",
    y = "Opioid-Related Mortality (per 10k)"
  )

# 2. Save the High-Resolution Artifact
# 'ggsave' preserves the 'real straight line' for publication
ggsave("Trend_combined.png", plot = p2, width = 6, height = 4.2, dpi = 300)

p2
```

## 3.4 Scatted plot with lines (combined and raw data): (Yearly) Relationship Between ED Visits and Mortality [Optional]
```{r}
# 1. Construct the Comparison Visualization
p_comparison <- ggplot(plot_data, aes(x = ER_annual_10k, y = Death_10k)) +
  
  # Layer 1: The 'Community Exposure' (Individual Points)
  # Keeping points monochromatic to focus the 'mind in motion' on the lines
  geom_point(size = 1, alpha = 0.15, color = "grey40") +
  
  # Layer 2: The 'Combined' Line (All data points)
  # This represents the total population (Observed + Imputed)
  geom_smooth(aes(color = "Combined", fill = "Combined"),
              method = "lm", 
              formula = y ~ x, 
              se = TRUE, 
              alpha = 0.1, 
              linewidth = 0.8) +
  
  # Layer 3: The 'Observed Only' Line
  # We filter the data within the aesthetic to show only 'Observed' rows
  geom_smooth(data = subset(plot_data, Status == "Observed"),
              aes(color = "Observed Only", fill = "Observed Only"),
              method = "lm", 
              formula = y ~ x, 
              se = TRUE, 
              alpha = 0.1, 
              linewidth = 0.8) +
  
  # Layer 4: High-Contrast Alignment
  # Using Charcoal for Observed and International Orange for Combined
  scale_color_manual(values = c("Combined" = "#D35400", "Observed Only" = "#2C3E50")) +
  scale_fill_manual(values = c("Combined" = "#D35400", "Observed Only" = "#2C3E50")) +
  
  # Standardizing the Interaction Lattice
  facet_wrap(~ year) + 
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Comparison of Observed and Combined Opioid Mortality Trajectories",
    x = "Emergency Department Visits (per 10k)",
    y = "Opioid-Related Mortality (per 10k)",
    color = "Trajectory Source",
    fill = "Trajectory Source"
  )

# 2. Save the High-Resolution Artifact
ggsave("obs_combined_comparison.png", plot = p_comparison, width = 6.5, height = 4.8, dpi = 300)

p_comparison
```



# 4. "Hot Spot" maps
## 4.1 [[[Figure 2]]]. Spatial Clustering of ED Visits and Methadone Distribution
### This Figure is newly produced in R. The older one in manuscript was created by ArcGIS.
```{r}
# # 1. Run predictions on original data
# predict1 <- predict(
#   fit0$fit,
#   newdata = fit0$imputed_df,
#   offset = log(fit0$imputed_df$pop_CDC)
# )
# 
# # Merge the predictions
# map_data <- left_join(counties_sf, predict1, by = "County")

study_counties <- df %>%
  mutate(County = str_to_title(County)) %>%
  distinct(County)

counties_study_sf <- counties_sf %>%
  mutate(County = str_to_title(County)) %>%
  semi_join(study_counties, by = "County")


spatial_summary <- df %>%
  mutate(
    County = str_to_title(County),
    ER_annual_percapita = as.numeric(ER_annual_percapita),
    METHADONE_annual_percapita = as.numeric(METHADONE_annual_percapita),
    ed_per_10k = ER_annual_percapita * 10000
  ) %>%
  group_by(County) %>%
  summarise(
    mean_ed_per_10k = mean(ed_per_10k, na.rm = TRUE),
    mean_methadone_percapita = mean(METHADONE_annual_percapita, na.rm = TRUE),
    .groups = "drop"
  )

map_fig2 <- counties_study_sf %>%
  left_join(spatial_summary, by = "County")


p_ed <- ggplot(map_fig2) +
  geom_sf(aes(fill = mean_ed_per_10k), color = "white", size = 0.3) +
  scale_fill_gradient(name = "ED visits\nper 10,000") +
  labs(title = "A. Opioid-related ED visits") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

p_methadone <- ggplot(map_fig2) +
  geom_sf(aes(fill = mean_methadone_percapita), color = "white", size = 0.3) +
  scale_fill_gradient(name = "Methadone\nper capita") +
  labs(title = "B. Methadone distribution") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

fig2 <- patchwork::wrap_plots(p_ed, p_methadone, ncol = 2) #+
  # patchwork::plot_annotation(
  #   title = "Figure 2. Spatial Distribution of ED Visits and Methadone Distribution",
  #   subtitle = "Study counties in Northwestern Alabama, mean values across 2016–2019"
  # )

fig2



```


## 4.2 Moran's I Summary Table [Optional]
```{r}
library(spdep)

# 1. Define the complete extraction function
# This function pulls all five metrics required for your Table 1 structure
extract_moran_metrics <- function(data_subset, var_name) {
  coords <- as.matrix(data_subset[, c("X", "Y")])
  nb <- knn2nb(knearneigh(coords, k = 4))
  lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
  
  # Execute the spatial test
  test_res <- moran.test(data_subset[[var_name]], lw, zero.policy = TRUE, na.action = na.omit)
  
  # Extract the exact columns to match the target layout
  data.frame(
    Morans_Index   = test_res$estimate[1],
    Expected_Index = test_res$estimate[2],
    Variance       = test_res$estimate[3],
    z_value        = test_res$statistic,
    p_value        = test_res$p.value
  )
}

# 2. Synthesize the data into the stacked 'Table 1' format
moran_stacked <- plot_data %>%
  group_by(year) %>%
  do({
    # Execute calculations for both variables
    m_er    <- extract_moran_metrics(., "ER_annual_10k")
    m_death <- extract_moran_metrics(., "Death_10k")
    
    # Assign the structural grouping labels
    m_er$Variable    <- "Emergency Department Visits"
    m_death$Variable <- "Opioid-Related Mortality"
    
    # Stack the rows vertically
    bind_rows(m_er, m_death)
  }) %>%
  ungroup()

# 3. Order and format the final manuscript table
table_1 <- moran_stacked %>%
  # Reorder columns to exactly match your image
  select(Variable, Year = year, Morans_Index, Expected_Index, Variance, z_value, p_value) %>%
  # Sort by Variable first, then Year, creating the grouped blocks
  arrange(desc(Variable), Year) %>%
  # Round to 4 decimal places for professional consistency
  mutate(across(where(is.numeric), ~ round(., 4)))



# Transpose the Data Architecture
table_1_transposed <- table_1 %>%
  # Melt the structural metrics down into a single 'Metric' column
  pivot_longer(
    cols = c(Morans_Index, Expected_Index, Variance, z_value, p_value),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  # Cast the temporal lattice (Years) across the top as columns
  pivot_wider(
    names_from = Year,
    values_from = Value
  ) %>%
  # Clean the administrative variable names for the final manuscript
  mutate(Metric = case_when(
    Metric == "Morans_Index"   ~ "Moran's I",
    Metric == "Expected_Index" ~ "Expected I",
    Metric == "Variance"       ~ "Variance",
    Metric == "z_value"        ~ "z-value",
    Metric == "p_value"        ~ "p-value"
  ))

table_1_transposed
# write.csv(table_1_transposed, "Moran_ER_Death.csv")
# 
# getwd()
```


# 5. Mediation Model
## 5.1 Imputation
```{r}
imputed_df <- fit0$fit$data
imputed_df$ER_annual <- imputed_df$ER_annual_percapita * imputed_df$pop_CDC

imputed_df$rural_factor <- as.factor(imputed_df$rural_factor)
imputed_df$urban <- as.factor(7 - as.numeric(imputed_df$rural_factor))
```

## 5.5 [Final] Mediator

### 5.5.0 Drugs Composite
```{r}
# Load necessary libraries
library(corrplot)

# Select and check correlations
opioid_cols <- c("OXYCODONE_annual_percapita_z", "HYDROCODONE_annual_percapita", 
                 "BUPRENORPHINE_annual_percapita_z", "METHADONE_annual_percapita_z")

cor_matrix <- cor(imputed_df[, opioid_cols], use = "complete.obs")

# Print matrix
print(round(cor_matrix, 2))



# 1. Select the three correlated variables
# We scale Hydrocodone as it was not z-scored in the original matrix
opioid_sub_data <- imputed_df %>%
  select(OXYCODONE_annual_percapita_z, 
         HYDROCODONE_annual_percapita, 
         BUPRENORPHINE_annual_percapita_z) %>%
  mutate(across(everything(), scale))

# 2. Run PCA to extract the primary 'Opioid Burden' signal
pca_opioid <- principal(opioid_sub_data, nfactors = 1)

# 3. Add the composite score to your dataframe
imputed_df$opioid_burden <- as.numeric(pca_opioid$scores)

# 4. Verify loadings - ensure they are positive (High score = High Burden)
print(pca_opioid$loadings)


```



### 5.5.0 SES Composite
```{r}

# 1. Select and Center/Scale (Z-scores)
ses_data <- imputed_df %>%
  select(gini_index, poverty_rate, noinsurance_rate) %>%
  mutate(across(everything(), scale))

# 2. Extract the First Principal Component
# In 'Classic Style', we assume SES is the primary latent factor
pca_result <- principal(ses_data, nfactors = 1)

# 3. Add the SES Score back to your dataset
# since indicators—poverty_rate, noinsurance_rate, and gini_index—all load positively on PC1 (meaning higher scores indicate more disparity). We need invert it.
imputed_df$ses_composite <- as.numeric(pca_result$scores) * -1
```





```{r}
# Stage 1: Predict ED Visits (M)
# Synthesizing OUD and SES to capture the non-linear temporal trajectory
form_mediator <- ER_annual ~
  (OXYCODONE_annual_percapita_z + BUPRENORPHINE_annual_percapita_z + HYDROCODONE_annual_percapita_z + METHADONE_annual_percapita_z) * (rural_factor + gini_index_z + poverty_rate_z + noinsurance_rate_z + time_index)


form_mediator <- ER_annual ~
  (OXYCODONE_annual_percapita_z + BUPRENORPHINE_annual_percapita_z + HYDROCODONE_annual_percapita_z + METHADONE_annual_percapita_z) * (rural_factor + gini_index_z + poverty_rate_z + noinsurance_rate_z + time_index + I(time_index^2))


# Stage 2: Predict Mortality (Y)
# Testing b-path moderation across an accelerating or decelerating timeline
# form_outcome <- Deaths ~
#   ER_annual_percapita_z * (rural_factor + ses_composite + time_index) +   opioid_burden + METHADONE_annual_percapita_z

# form_outcome <- Deaths ~
#   OXYCODONE_annual_percapita_z + BUPRENORPHINE_annual_percapita_z + HYDROCODONE_annual_percapita_z + METHADONE_annual_percapita_z + ER_annual_percapita_z * (rural_factor + gini_index_z + poverty_rate_z + noinsurance_rate_z + time_index)
  

form_outcome <- Deaths ~
  OXYCODONE_annual_percapita_z + BUPRENORPHINE_annual_percapita_z + HYDROCODONE_annual_percapita_z + METHADONE_annual_percapita_z + ER_annual_percapita_z * (rural_factor + gini_index_z + poverty_rate_z + noinsurance_rate_z + time_index + I(time_index^2))

```



```{r}
fit_mediator <- sdmTMB(
  formula = form_mediator,
  data    = imputed_df,
  mesh    = mesh,
  family  = nbinom2(link = "log"), # Changed from poisson
  #spatial = "on",
  spatial = "off",
  offset  = "log_pop",
  priors  = sdmTMBpriors(matern_s = pc_matern(range_gt = 10, sigma_lt = 2))
)


# Fit the outcome model similarly
fit_outcome <- sdmTMB(
  formula = form_outcome,
  data    = imputed_df,
  mesh    = mesh,
  family  = nbinom2(link = "log"), # Changed from poisson
  #spatial = "on",
  spatial = "off",
  offset  = "log_pop",
  priors  = sdmTMBpriors(matern_s = pc_matern(range_gt = 10, sigma_lt = 2))
)


```



### 5.5.1 [[[Table 3]]]. Medication-Specific Associations from Spatiotemporal Model
```{r}
# 1. Extract the fixed effects for each model
# Using conf.int = TRUE will also calculate the 95% CIs if you need them later
tidy_med <- tidy(fit_mediator, conf.int = TRUE) %>%
  mutate(Model = "Mediator (Rural Base)")

# tidy_med_urb <- tidy(fit_mediator_urban, conf.int = TRUE) %>%
#   mutate(Model = "Mediator (Urban Base)")

tidy_out <- tidy(fit_outcome, conf.int = TRUE) %>%
  mutate(Model = "Outcome (Mortality)")

# 2. Combine the rows into one dataset
#all_models <- bind_rows(tidy_med, tidy_med_urb, tidy_out)
all_models <- bind_rows(tidy_med, tidy_out)

# 3. Format the estimates and standard errors (independent of stars)
formatted_models <- all_models %>%
  mutate(
    # Calculate the z-statistic and p-value
    z_val = estimate / std.error,
    p_val = 2 * pnorm(-abs(z_val)),
    
    # Assign stars based on the p-value
    stars = case_when(
      p_val < 0.001 ~ "***",
      p_val < 0.01  ~ "**",
      p_val < 0.05  ~ "*",
      p_val < 0.1   ~ ".",
      TRUE          ~ ""
    ),
    
    # Format numbers ONLY: e.g., "-0.08 (0.25)"
    Result = sprintf("%.2f (%.2f)", estimate, std.error)
  ) %>%
  dplyr::select(term, Model, Result, stars)

# 4. Pivot the table to create paired columns for Result and stars per model
final_table <- formatted_models %>%
  pivot_wider(
    names_from = Model,
    values_from = c(Result, stars),
    names_glue = "{Model}_{.value}", # Automatically names columns (e.g., "BaseModel_Result", "BaseModel_stars")
    values_fill = list(Result = "-", stars = "") # Assigns appropriate blank values for missing data
  )


# 5. Print the table
kable(final_table, align = "l")

# Optional: You can add a note at the bottom of your markdown or CSV 
# explaining the legend: *** p < 0.001, ** p < 0.01, * p < 0.05,. p < 0.1

# Save the table to your current working directory
#write.csv(final_table, "moderated_mediation_results.csv", row.names = FALSE)
#write.csv(final_table, "moderated_mediation_results_quadratic.csv", row.names = FALSE)
```




### 5.5.2 Adjusting for Distributional Assumptions (Overdispersion)
```{r}
# 2. Check Assumptions using DHARMa residuals
library(DHARMa)
# Simulate from the joint precision matrix to test residuals
sim_res_med <- simulate(fit_mediator, nsim = 500, type = "mle-mvn")
dharma_residuals(sim_res_med, fit_mediator)
```


# 6. Shiny

```{r}
# 4. Save this object specifically for the Shiny App
save(final_fit, df, file = "best_model.RData")

load(paste0(repos,"/Data/best_model.RData"))

```














